Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-01 manu-draft.qmd
finish discussion text but need to add refs

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments. At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both occurred showed consistent increases across the period of record. Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.

@@

The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area [@lizcano-sandoval2022]. Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south. These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay. Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist. Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive. For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth. These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times, whereas the effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present. Competitive differences between seagrasses and macroalgae are poorly understood in these systems, in addition to insufficient data to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting macroalgae growth and interactios with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing esearch has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats. In southwest Florida, the most common example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.

@@

This study provided a detailed assessment of long-term water temperature and salinity changes in Tampa Bay supported by datasets from three long-term monitoring programs of different length and sampling design. An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that TampaBay has become hotter and fresher with the trends likely continuing in the future. Simple regression

Case for continuous monitoring

OTB is likely at a tipping point...

Manamagement implication - we can't do much to control temp and salinity, but we can temper expectations for how to manage the bay. Adaptive capacity/resiliency is likely decreasing because of these "co-morbidities", so we have to temper expectations for other stressors that we can control.

>>>>

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments. At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both crossed thresholds showed consistent increases across the period of record. Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.

@@

The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area [@lizcano-sandoval2022]. Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south. These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay. Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist. Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive. For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth. These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times. The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present. Competitive differences between seagrasses and macroalgae are poorly understood in these systems, in addition to insufficient data to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats. In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.

@@

This study provided a detailed assessment of long-term water temperature and salinity changes in Tampa Bay supported by datasets from three long-term monitoring programs of different length and sampling design. An evaluation of each dataset showed a clear pattern of increasing temperature and decreasing salinity mirrored by long-term changes in air temperature and precipitation, suggesting that Tampa Bay has become hotter and fresher with the trends likely continuing in the future. Simple regression models provided weak, but partially-supporting evidence that these changes can be linked to recent seagrass losses since 2016.  Our models suggested that rising temperatures and decreasing salinities are additively affecting seagrasses, rather than multiplicatively as evidenced by lack of significant interactions of these stressors.  Future analyses may show more significant associations between physical habitat conditions and seagrass change as the trends are very likely to continue to push seagrass further outside of their tolerance ranges. These analyses should be supported by additional data, particularly high-resolution continuous monitoring data that provide a more precise assessment of stress at multiple time-scales.

Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at supporting ecological resources in Tampa Bay. Management actions that have historically been effective may have degraded capacity to support ecosystem status due to changing resilience of ecological resources with climate change.  For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay.  As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated.  Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modification).  These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where trends have been most pronounced.  Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded systems due to hysteresis.
